---
title: "Effect of weather events on US health and Economy, 2002 to 2011"
author: "Gill Mundin"
date: "15 July 2019"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##Synopsis
Weather data were used to identify the events that had most impact on health and economy.  Data were available from 1950, but results from the period of 2002 to 2011 were reported as the data were more complete.

There were multiple classifications of the same type of weather event.  10 broader groups were chosen and the events were reclassified according to these broad groups to create the analysis datasets.

Using the broad groups, in the period between 2002 and 2011, HEAT was responsible for the most deaths (>2000) and TORNADO was responsible for the most injuries (>10,000).

Flood was most damaging to property and crops, causing > $140 billion and $4 billion respectively.  Hurricane was the next most damaging for property and crops.


##Data processing
The sections below describe how the data were loaded and processed for analysis

###Packages used
```{r packages}
library(dplyr)
library(lattice)
library(tidyr)

```

###Load data

A link to the storm data is found on the Coursera website:
https://www.coursera.org/learn/reproducible-research/peer/OMZ37/course-project-2

Right click the Storm Data hyperlink and copy link address:
https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

Assign the link address to variable url, and use to download file to working directory.

Use the chunk option cache = TRUE to save knit time.

```{r load_data, cache = TRUE}

url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
filename <- "storm-data.csv.bz2"  # Saved file name, saved to working directory
download.file(url, filename)
data <- read.csv(filename)

```


###Process data for analysis
The EVTYPE variable in the data has many similar events which need to be grouped.
The grouping will not be perfect but demonstrates an approach that might be improved upon.
```{r group_events}

data$group <- NA #Adds a column of NAs with colname group to the data dataframe

snowrows <- grep("snow",
                 data$EVTYPE,
                 ignore.case = TRUE)

icerows <- grep("ice", data$EVTYPE, ignore.case = TRUE)
floodrows <- grep("flood", data$EVTYPE, ignore.case = TRUE)
hailrows <- grep("hail", data$EVTYPE, ignore.case = TRUE)
tornadorows <- grep("tornado", data$EVTYPE, ignore.case = TRUE)
firerows <- grep("fire", data$EVTYPE, ignore.case = TRUE)
windrows <- grep("wind", data$EVTYPE, ignore.case = TRUE)
heatrows <- grep("heat", data$EVTYPE, ignore.case = TRUE)
coldrows <- grep("cold", data$EVTYPE, ignore.case = TRUE)
hurricanerows <- grep("hurricane", data$EVTYPE, ignore.case = TRUE)


# Replace the NAs in the group column with one of the weather event groups
data[snowrows, "group"] <- "SNOW"
data[icerows, "group"] <- "ICE"
data[floodrows, "group"] <- "FLOOD"
data[hailrows, "group"] <- "HAIL"
data[tornadorows, "group"] <- "TORNADO"
data[firerows, "group"] <- "FIRE"
data[windrows, "group"] <- "WIND"
data[heatrows, "group"] <- "HEAT"
data[coldrows, "group"] <- "COLD"
data[hurricanerows, "group"] <- "HURRICANE"
```


Generate an analysis dataset for health question. 
Relevant variables are STATE, group, FATALITIES, INJURIES

Generate an analysis dataset for economy question. 
Relevant variables are BGN_DATE, BGN_TIME, STATE, group, PROPDMG:CROPDMGEXP, REMARKS, REFNUM



```{r data_sets}
health <- data %>% select(STATE, group, FATALITIES, INJURIES)
tidyhealth <- gather(health, outcome, count, -STATE, -group)
tidyhealth$outcome <- as.factor(tidyhealth$outcome)

# create subsets for last 10 years, 
# convert data column to date

data$BGN_DATE <- as.Date(as.POSIXct(data$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"))

date1 <- as.Date("2001-12-31")
date2 <- as.Date("2012-01-01")

data00s <- data[data$BGN_DATE > date1 & data$BGN_DATE < date2, ]
h00s <- data00s %>% select(STATE, group, FATALITIES, INJURIES)
tidy00s <- gather(h00s, outcome, count, -STATE, -group)
tidy00s$outcome <- as.factor(tidy00s$outcome)

econ <- data %>% select(BGN_DATE, BGN_TIME, STATE, group, PROPDMG:CROPDMGEXP, REFNUM)
econ00s <- data00s %>% select(BGN_DATE, BGN_TIME, STATE, group, PROPDMG:CROPDMGEXP, REFNUM)

```




###Determine which weather event caused most injuries and fatalities across entire US

Use tapply to determine the number of injuries and fatalities for each weather event using the assigned groups rather than the recorded events

Entire data set and data limited to the most recent 10 years were examined.

```{r health, cache=FALSE}
harm_event <- with(tidyhealth,
                   tapply(count, 
                          list(group, outcome), 
                          sum,
                          na.rm = TRUE)) %>% 
        as.data.frame()

harm_event00s <- with(tidy00s,
                      tapply(count, 
                             list(group, outcome), 
                             sum,
                             na.rm = TRUE)) %>% 
        as.data.frame()

fatal <- harm_event[which.max(harm_event$FATALITIES), ] # harm_event must be df for this to work
fatalevent <- rownames(fatal)
nfatal <- fatal[1,1]

fatal00s <- harm_event00s[which.max(harm_event00s$FATALITIES), ] # harm_event must be df for this to work
fatalevent00s <- rownames(fatal00s)
nfatal00s <- fatal00s[1,1]

injury <- harm_event[which.max(harm_event$INJURIES), ]
injuryevent <- rownames(injury)
ninjury <- injury[1,2]

injury00s <- harm_event00s[which.max(harm_event00s$INJURIES), ]
injuryevent00s <- rownames(injury00s)
ninjury00s <- injury00s[1,2]
```

Rank the data health data and determine the top 5 events most damaging to health.

```{r rank_health}
fatal_rank <- harm_event[order(harm_event$FATALITIES, decreasing = TRUE), ]
fatal_top_5 <- fatal_rank[1:5, ]

fatal00s_rank <- harm_event00s[order(harm_event00s$FATALITIES, decreasing = TRUE), ]
fatal00s_top_5 <- fatal00s_rank[1:5, ]

injury_rank <- harm_event[order(harm_event$INJURIES, decreasing = TRUE), ]
injury_top_5 <- injury_rank[1:5, ]

injury00s_rank <- harm_event00s[order(harm_event00s$INJURIES, decreasing = TRUE), ]
injury00s_top_5 <- injury00s_rank[1:5, ]

fatal_top_5
injury_top_5

fatal00s_top_5
injury00s_top_5

```

###Use the EXP information to calculate actual damage to property - **limited to latest decade**
Based on this internet article https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html
create an EXP dataframe which matches entires to multiplication factors

```{r exp_dataframe}
expdf <- as.data.frame(unique(econ$PROPDMGEXP))
colnames(expdf) <- "PROPDMGEXP"
expdf$factor = c(10^3, 10^6, 1, 10^9, 10^6, 1, 10, 10, 10, 1, 10, 10, 10, 100, 10, 100, 1, 10, 10)


# Merge expdf with econ
econ1 <- merge(econ00s, expdf, by = "PROPDMGEXP")
econ2 <- merge(econ1, expdf, by.x = "CROPDMGEXP", by.y = "PROPDMGEXP")


# Calculate numerical value for property damage using mutate

econvalue <- econ2 %>%  mutate(property = PROPDMG * factor.x, crop = CROPDMG * factor.y)

# Now sum across all the event types

propsum <- with(econvalue,
                tapply(property,
                       group,
                       sum,
                       na.rm = T)) %>% 
        as.data.frame()
colnames(propsum) <- "propvalue"
propsum$event <- rownames(propsum)

cropsum <- with(econvalue,
                tapply(crop, 
                       group, 
                       sum,
                       na.rm = T)) %>% 
        as.data.frame()
colnames(cropsum) <- "cropvalue"
cropsum$event <- rownames(cropsum)

propmax <- propsum[which.max(propsum$propvalue), ]
propmax

cropmax <- cropsum[which.max(cropsum$cropvalue), ]
cropmax


prop00s <- propsum[which.max(propsum$propvalue), ] # propsum must be df for this to work
propevent00s <- rownames(prop00s)
nprop00s <- prop00s[1,1]

crop00s <- cropsum[which.max(cropsum$propvalue), ] # cropsum must be df for this to work
cropevent00s <- rownames(crop00s)
ncrop00s <- crop00s[1,1]

```

```{r rank_economy}
prop_rank <- propsum[order(propsum$propvalue, decreasing = TRUE), ]

crop_rank <- cropsum[order(cropsum$cropvalue, decreasing = TRUE), ]  

prop_rank$propvalbil <- prop_rank$propvalue / 1000000000
crop_rank$cropvalbil <- crop_rank$cropvalue / 1000000000

prop_rank
crop_rank

```


##Results
####Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?

`r fatalevent` was the weather event responsible for the most injuries and fatalities across the United States both for the entire data set and the most recent 10 year period:

```{r}
fatal
injury

fatal00s
injury00s

```

`r fatalevent` was responsible for `r nfatal` fatalities and `r as.integer(ninjury)` injuries in the period studied.


`r fatalevent00s` was responsible for `r nfatal00s` fatalities and `r as.integer(ninjury00s)` injuries in the period between 2002 and 2011.


```{r healthplots, fig.height=9, cache= FALSE}

par(mfrow=c(2, 1), 
    oma = c(0, 0, 2, 0),
    mar = c(12, 8, 2, 2),
    mgp = c(4, 1, 0))


barplot(fatal_top_5$FATALITIES, 
                     names.arg = rownames(fatal_top_5), 
                     las = 2,
                     col = "light blue",
                     main = "Deaths",
                     ylab = "# deaths")

barplot(injury_top_5$INJURIES, 
                     names.arg = rownames(injury_top_5), 
                     las = 2,
                     col = "orange",
                     main = "Injuries",
                     ylab = "# injuries")

mtext("Top 5 events for deaths and injuries over period studied", 
      outer = TRUE, 
      cex = 1.5)

```


```{r healthplots00s, fig.height=9, cache= FALSE}

par(mfrow=c(2, 1), 
    oma = c(0, 0, 2, 0),
    mar = c(12, 8, 2, 2),
    mgp = c(4, 1, 0))


barplot(fatal00s_top_5$FATALITIES, 
        names.arg = rownames(fatal00s_top_5), 
        las = 2,
        col = "light blue",
        main = "Deaths",
        ylab = "# deaths")

barplot(injury00s_top_5$INJURIES, 
        names.arg = rownames(injury00s_top_5), 
        las = 2,
        col = "orange",
        main = "Injuries",
        ylab = "# injuries")

mtext("Top 5 events for deaths and injuries 2002 - 2011", 
      outer = TRUE, 
      cex = 1.5)

```



The panel plot shows that Tornado was the most significant factor in deaths and injuries.  Excessive heat caused the next greatest amount of deaths.  All other weather events caused far few deaths or injuries than tornado.

###Across the United States, which types of events have the greatest economic consequences?

This analysis was limited to the period between 2002 and 2011.

`r propevent00s` caused the most damage to property, and `r cropevent00s` the most damage to crops during the period examined.
The next most damaging weather event to property and crops was HURRICANE.


```{r economy_plots, fig.height=9}

par(mfrow=c(2, 1), 
    oma = c(0, 0, 2, 0),
    mar = c(12, 8, 2, 2),
    mgp = c(4, 1, 0))

barplot(prop_rank$propvalbil, 
        names.arg = prop_rank$event, 
        las = 2,
        col = "light blue",
        main = "Property damage",
        ylab = "Value (billion $)")

barplot(crop_rank$cropvalbil, 
        names.arg = crop_rank$event, 
        las = 2,
        col = "orange",
        main = "Crop damage",
        ylab = "Value (billion $)")

mtext("Top 10 weather events property and crop damage 2002 - 2011", 
      outer = TRUE, 
      cex = 1.5)        
        
        